### 2022-12-20 introduction/introduction.tex
Added to outline of protein sequence-structure relationships and refined wording on "biological role"

%    B. Definition and discussion of protein disorder
%        1. Structural heterogeneity, conformational ensembles, floppy
%        2. Major sequence characteristics and structural "modes"?
%    C. Disordered proteins are challenging to study with traditional methods
%    D. Genome sequencing and computational tools revealed disordered proteins are ubiquitous
%IV. Disordered proteins perform varied functions

@@

%V. Disordered proteins are enriched in transcriptional machinery

@@

%VI. Disordered proteins are associated with phase separation

@@

%VII. Phase separation via IDRs is a possible mechanism for initiating transcription

@@

%Discuss more recent developments in field and propose driving questions

@@

Clearly, understanding the relationship between the sequence, structure, and function of proteins is essential for unraveling more complex biological phenomena. While biologists are generally most interested in the function of a protein since it is the most directly associated with its biological role\footnote{Function and biological role are frequently, though not always, exact synonyms since biological role is often defined more broadly and loosely than function. For example, the strict function of hemoglobin is to bind oxygen, an activity it shares with a related protein myoglobin. However, the two have different roles. Hemoglobin is found in red blood cells where it acts as a carrier during transport. In contrast, myoglobin is found in muscle cells, where it stores oxygen until needed.}, function and biological role are not always easily identified or measured. Thus, determining a protein's structure is frequently the first step of more detailed studies of its function. Though in recent years researchers have developed powerful computational tools that can accurately predict structure from sequence alone, historically structures were determined experimentally, and experimental methods still remain the gold standard. While many methods can reveal information about the structure of a protein, the most powerful techniques, X-ray crystallography, NMR spectroscopy, and cryogenic electron microscopy (cryo-EM), can map the spatial coordinates of every atom in a protein. However, this resolution requires extremely pure samples of the protein of interest. Since proteins are only produced by living systems, preparations begin with a complex mixture consisting of cells or tissue, and the protein of interest is isolated with series of extraction and purification steps. Some proteins are only produced in small amounts or degrade easily, so each step may require substantial optimization to achieve a sufficient yield. When structural techniques were first developed in the late 1950s, they were so time-consuming that a graduate student could dedicate an entire PhD to solving a single protein structure. Many developments have substantially accelerated the process, but it remains a labor-intensive technique which may require several months of effort. However, the result is a powerful map that scientists use to suggest hypotheses and interpret data.

The success of structural methods at elucidating the molecular details of protein function cemented the view that function largely depends on the presence of a fixed structure. While scientists understood proteins were not completely rigid and could adopt a variety of related structures or ``conformations,'' many believed that functional proteins largely had a single dominant structure. Despite its strength, exceptions to the structure-function paradigm were known. For example, elastin is a protein secreted by cells which allows tissues like skin or blood vessels to repeatedly expand and contract. It imparts this elasticity by forming networks of disordered chains that act like molecular springs. When a tissue experiences a force, the chains stretch to accommodate it. When the force is removed, the chains return to their random orientations, which reduces their end-to-end length and forces the tissue to return to its original shape~\cite{Vrhovski1998, Alberts2014}. However, as a result of this unique role in providing tissue elasticity, elastin's disorder was viewed as a specific adaptation rather than a general mechanism of protein function. In other cases, proteins had regions which returned undefined or highly variable atomic coordinates when analysed with structural techniques, but because these regions were often short loops between structured regions, they were seen as ``linkers'' which facilitated the structure of the functional portions of the protein. By the early 2000s, though, enough exceptions had accumulated that scientists began to recognize that fully and partially disordered proteins were involved in many biological processes~\cite{Plaxco1997, Wright1999, Dunker2001}. Many of these examples were proteins which folded on binding to their targets, commonly other proteins. This mechanism was a departure from the prevailing model of interactions between biological macromolecules which required highly stable and complementary interfaces, like two puzzle pieces fitting together. As a result, scientists speculated that disorder was an adaptation that allowed proteins to efficiently relay and regulate signals by enabling interactions with many possible targets. Furthermore, the flexibility of disordered protein would permit environmental conditions to easily modulate these interactions. In the following years, as the complete genomes of several scientifically important model organisms such as \textit{S. cerevisiae} (baker's yeast), \textit{C. elegans} (roundworm), and \textit{D. melanogaster} (fruit fly) were sequenced for the first time, researchers applied computational methods for predicting disorder to the proteins inferred from their genomes. They discovered that disorder is ubiquitous in eukaryotic\footnote{All life belongs to one of three categories, or domains. Two, Archaea and Bacteria, are all single-celled organisms with simple cellular structures. In contrast, the cells of members of Eukarya, \textit{i.e.}, eukaryotes, are complex and contain substructures called organelles, among other differences. All animals, plants, and fungi are eukaryotes, though the domain includes many microorganisms as well.} organisms, with estimates of the fraction of proteins containing disordered segments of greater than 30 residues\footnote{The amino acids that compose the links of a protein chain are conventionally called residues to distinguish them from their related, but chemically distinct, free forms.} ranging between 28 and 63\%~\cite{Dunker2000, Ward2004}. For reference, although the lengths of proteins can vary dramatically, a ``typical'' protein contains on the order of a few hundred residues, so these regions can compose a significant fraction of a protein's length. Thus, disorder was recognized as a pervasive but poorly understood feature of proteins.

>>>>

%    B. Disordered proteins are challenging to study with traditional methods
%    C. Genome sequencing and computational tools revealed disordered proteins are ubiquitous
%IV. Sequence-structure relationships in protein disorder
%    A. Prompted by these observations, researchers began to clarify the sequence-structure-function relationships that characterize protein disorder
%        1. They discovered that related to structure and sequence but the details differ
%    B. Though structured proteins are not completely rigid, they generally have a small number of structures
%        1. As mentioned previously, even before disorder was recognized as a common feature of proteins, scientists understood proteins were not completely rigid
%        2. At the molecular level, everything is in constant motion. For example, at room temperature the average water molecule moves at a speed of SPEED.
%            i. However, collisions and interactions with other water molecules generally prevent them from escaping into the atmosphere
%        3. Proteins are also in motion, but they are constrained both by the rigid bonds and interactions between amino acid residues in the chain
%        4. Thus, all proteins flex and breathe. However, these motions are minor variations on the stable structure.
%        5. In other cases, proteins have multiple structures which are related to different functional states.
%            i. For example, hemoglobin has two forms, traditionally called T and R.
%            ii. The T form is hemoglobin's oxygen-free state.
%            iii. On binding oxygen, its structure changes slightly, with the shift on the scale of a few hydrogen atoms at the points of greatest change.
%            iv. This changes the orientation between the atoms in hemoglobin that interact with oxygen, allowing it to bind more tightly.
%            v. Promotes oxygen uptake at other three binding sites, promoting efficiently transport
%        6. Although proteins have multiple conformations, they are generally stable and the intermediates are not.
%        7. Thus, proteins with multiple conformations generally exist in discrete states.
%    C. Protein disorder means their atoms do not have fixed spatial relationships
%       1. In contrast, disordered proteins are described as ``conformational ensembles''
%       2. They are constantly sampling conformations which may have large differences in the spatial relationships between residues
%    D. Protein folding driven by hydrophobic interactions
%       1. Disordered regions largely characterized by enrichment and depletion in hydrophilic and hydrophobic residues, respectively
%       2. Hydrophobic residues are ``water-fearing,'' and like oil in water, they tend to self-associate.
%       3. Hydrophilic residues are ``water-loving,'' and easily interact with water.
%       4. Hydrophobic interacts are a major force driving protein folding, and many folded proteins have a hydrophobic core
%    E. Sequence-structure relationships in disordered proteins
%       1. Although, disordered regions are conformationally heterogeneous, they
%V. Functional roles of protein disorder

@@

%VI. Disordered proteins are enriched in transcriptional machinery

@@

%VII. Disordered proteins are associated with phase separation

@@

%VIII. Phase separation via IDRs is a possible mechanism for initiating transcription

@@

@@

Clearly, understanding the relationship between the sequence, structure, and function of proteins is essential for unraveling more complex biological phenomena. Though biologists study all three properties of proteins, they are generally most interested in function since it is the most directly related to the biological processes the protein takes part in\footnote{Function generally refers to \textit{molecular} function which is a description of a specific chemical activity possessed by a protein. A biological process, however, is the larger ``biological program'' which is accomplished by the action of multiple linked molecular functions. For example, the molecular function of hemoglobin is to bind oxygen, an activity it shares with a related protein myoglobin. However, the two have different roles in the process of oxygen transport and storage. Hemoglobin is found in red blood cells where it acts as a carrier during transport. In contrast, myoglobin is found in muscle cells, where it stores oxygen until needed.}, function and biological process are not always easily identified or measured. Thus, determining a protein's structure is frequently the first step of detailed studies of its function. Though in recent years researchers have developed powerful computational tools that can accurately predict structure from sequence alone, historically structures were determined experimentally, and experimental methods still remain the gold standard. While many methods can reveal information about the structure of a protein, the most powerful techniques, X-ray crystallography, NMR spectroscopy, and cryogenic electron microscopy (cryo-EM), can map the spatial coordinates of every atom in a protein. However, this resolution requires extremely pure samples of the protein of interest. Since proteins are only produced by living systems, preparations begin with a complex mixture consisting of cells or tissue, and the protein of interest is isolated with series of extraction and purification steps. Some proteins are only produced in small amounts or degrade easily, so each step may require substantial optimization to achieve a sufficient yield. When structural techniques were first developed in the late 1950s, they were so time-consuming that a graduate student could dedicate an entire PhD to solving a single protein structure. Many developments have substantially accelerated the process, but it remains a labor-intensive technique which may require several months of effort. However, the result is a powerful map that scientists use to suggest hypotheses and interpret data.

The success of structural methods at elucidating the molecular details of protein function cemented the view that function largely depends on the presence of a fixed structure. While scientists understood proteins were not completely rigid and could adopt a variety of related structures, many believed that functional proteins largely had a single dominant structure. Despite its strength, exceptions to the structure-function paradigm were known. For example, elastin is a protein secreted by cells which allows tissues like skin or blood vessels to repeatedly expand and contract. It imparts this elasticity by forming networks of disordered chains that act like molecular springs. When a tissue experiences a force, the chains stretch to accommodate it. When the force is removed, the chains return to their random orientations, which reduces their end-to-end length and forces the tissue to return to its original shape~\cite{Vrhovski1998, Alberts2014}. However, as a result of this unique role in providing tissue elasticity, elastin's disorder was viewed as a specific adaptation rather than a general mechanism of protein function. In other cases, proteins had regions which returned undefined or highly variable atomic coordinates when analysed with structural techniques, but because these regions were often short loops between structured regions, they were seen as ``linkers'' which facilitated the structure of the functional portions of the protein. By the early 2000s, though, enough exceptions had accumulated that scientists began to recognize that fully and partially disordered proteins were involved in many biological processes~\cite{Plaxco1997, Wright1999, Dunker2001}. Many of these examples were proteins which folded on binding to their targets, commonly other proteins. This mechanism was a departure from the prevailing model of interactions between biological macromolecules which required highly stable and complementary interfaces, like two puzzle pieces fitting together. As a result, scientists speculated that disorder was an adaptation that allowed proteins to efficiently relay and regulate signals by enabling interactions with many possible targets. Furthermore, the flexibility of disordered protein would permit environmental conditions to easily modulate these interactions. In the following years, as the complete genomes of several scientifically important model organisms such as \textit{S. cerevisiae} (baker's yeast), \textit{C. elegans} (roundworm), and \textit{D. melanogaster} (fruit fly) were sequenced for the first time, researchers applied computational methods for predicting disorder to the proteins inferred from their genomes. They discovered that disorder is ubiquitous in eukaryotic\footnote{All life belongs to one of three categories, or domains. Two, Archaea and Bacteria, are all single-celled organisms with simple cellular structures. In contrast, the cells of members of Eukarya, \textit{i.e.}, eukaryotes, are complex and contain substructures called organelles, among other differences. All animals, plants, and fungi are eukaryotes, though the domain includes many microorganisms as well.} organisms, with estimates of the fraction of proteins containing disordered segments of greater than 30 residues\footnote{The amino acids that compose the links of a protein chain are conventionally called residues to distinguish them from their related, but chemically distinct, free forms.} ranging between 28 and 63\%~\cite{Dunker2000, Ward2004}. For reference, although the lengths of proteins can vary dramatically, a ``typical'' protein contains on the order of a few hundred residues, so these regions can compose a significant fraction of a protein's length. Thus, disorder was recognized as a pervasive but poorly understood feature of proteins.
